ranacapa: An R package to explore environmental DNA data with exploratory statistics and interactive visualizations

### 2018-10-04 manuscript.Rmd
version for feedback from ec zjg mcc

---
title: 'ranacapa: An R package to explore environmental DNA data with exploratory statistics and interactive visualizations'
author: Gaurav S. Kandlikar (1,+), Zachary J. Gold (1), Madeline C. Cowen (1), Rachel  S. Meyer (1), Amanda C. Freise (2), Nathan J.B. Kraft (1), Jordan Moberg-Parker (2), Joshua Sprague (3), David Kushner (3), and Emily E. Curd (1)
fontsize: 11pt
output: 
  word_document:
    reference_docx: rendering-templates/word_template.docx
urlcolor: blue
always_allow_html: yes
bibliography: bibliography.bib
csl: rendering-templates/ecology-letters.csl
---

@@

The targeted amplification and sequencing of DNA that living organisms shed into their physical environment, termed "environmental DNA metabarcoding" or "eDNA sequencing",  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].  Sequencing of environmental DNA extracted from field-collected soil, water, or sediment samples can shed light on a range of questions, ranging from profiling the composition of ancient plant and animal communities [@Pedersen_2014], to motoring populations of rare or endangered species [@Balasingham_2017]. As the cost of eDNA sequencing declines and sample collection techniques become more streamlined (e.g. @Thomas_2018), professional research scientists are increasingly using eDNA sequencing as a platform to partner with a diversity of community partners, including natural resource managers, undergraduate students, and citizen scientists in primary research. However, developing robust and impactful community science programs that engage community partners in all steps of the research process remains a challenge.

eDNA sequencing-based projects work well for community science partnerships because non-experts can be quickly trained to collect samples in the field, and because eDNA sequencing is an exciting framework for research pertinent to disciplines such as medicine, agriculture, ecology, and geography [@Deiner_2017]. Community partners in such programs can have heterogeneous backgrounds, ranging from curious members of the public for whom collecting samples in the field is the first scientific research experience (e.g. University of California’s CALeDNA program, http://www.ucedna.com/), to professional natural resource managers who regularly collaborate with research scientists (e.g. Center for Ocean Solutions’ eDNA project, https://oceansolutions.stanford.edu/project-environmental-dna). A key ingredient to ensure sustained engagement of community partners is that they should be able to participate across multiple stages of the research project, not only in sample collection [@Pandya_2012; @ECSA_2015]. This can be a challenge for eDNA sequencing-based community science programs because although it is relatively easy to train community partners to collect eDNA samples, it is far more challenging to train them to interact with and visualize the results from these studies. Indeed, learning the bioinformatic tools necessary for visualizing and analyzing the large, multidimensional datasets generated in these studies can be difficult for professional researchers (Carey & Papin 2018), let alone for the non-technical audience of community science programs.

To address this challenge, we created an R package “ranacapa”, at the core of which is a Shiny webapp with which users can visualize and perform simple community ecology analyses on results from eDNA sequencing studies. `ranacapa` complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, `ranacapa` includes brief explanations and links to additional educational resources to provide users with an overview of basic data analyses used in eDNA studies. `ranacapa` works with community matrices generated via QIIME () or the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), which is used extensively by the CALeDNA program.

In the remainder of this manuscript, we describe `ranacapa` and demonstrate its use by two community science partnerships based at the University of California, Los Angeles (UCLA): first, a collaboration between eDNA researchers and resource managers at the National Park Service, and second, a partnership between community ecology researchers and an undergraduate microbiology course at UCLA. As we show in the Use Cases, empowering community partners to interact with the data and perform simple but insightful community ecology analyses can help make these collaborations more enriching and valuable to both parties.

## Implementation
At the core of `ranacapa` is a Shiny webapp (@Chang_2018), which is avialable at http://gauravsk.shinyapps.io/ranacapa or with `ranacapa::runRanacapa()`. The package also includes two categories of helper functions (Table 1) that transform user-uploaded taxonomy and metadata tables into `R` objects that can be visualized and analuzed using Phyloseq [@McMurdie_2013] and Vegan [@Oksanen_2018]. `ranacapa` is available for installation on Github or CRAN:

@@

- **Sequencing depth**:  Introduces the potential for variation in sequencing depth among samples and explains the basic logic behind rarefying samples in metagenomics studies. Allows users to rarefy the dataset to a sampling depth (Figure 1). The documentation on this tab also acknowledges recent disagreement regarding the value of rarefying in metabarcoding and eDNA sequencing studies [@McMurdie_2014], and allows users to proceed through the rest of the app without rarefying samples.

- **Taxonomy heatmap**: Shows the taxonomy matrix as an interactive heatmap made using `heatmaply::heatmaply()`, where the color of each cell represents the number of times a given taxon was sequenced in a sample (Figure 2).

@@

- **Alpha diversity plots**: Introduces the concept of Alpha diversity as the local diversity measured in a single habitat or sample. Users can plot Alpha diversity as observed taxon richness or as Shannon Diversity per sample, or can group samples according to a variable in the metadata file (Figure 4).

- **Alpha diversity statistics**: Allows users to choose a variable from the metadata, and generates an Alpha diversity ANOVA table according to the user-selected variable. The tab also shows the output from  a post-hoc Tukey test.

- **Beta diversity plots**: Introduces the concept of Beta diversity as the turnover in species composition across habitats (or samples). The tab includes an ordination plot generated by `phyloseq::plot_ordination()` based on an ordination made with `phyloseq::ordinate(., method = "PCoA")`. Points on the PCoA plot are colored according to a user-selected metadata variable (Figure 5). The tab also shows dissimilarity among sites according in a dendrogram generated with Ward’s cluster analysis (`stats::hclust(distance_object, method = "ward.d2")`), where `distance_object` is made using `phyloseq::distance()`. For both figures, users can toggle between using Jaccard and Bray-Curtis dissimilarity.

- **Beta diversity statistics**: Shows results from two statistical tests of species turnover across site: first, a multivariate ANOVA implemented with `vegan::adonis())`, which shows users results from a statistical test of taxon turnover across sites, and second, a statistical test of heterogeneity of variances among samples implemented with `vegan::betadisper()`, which shows results from a statistical test that compares the degree of sample-to-sample variation within habitats (or within other user-selected groups).

@@

### Input file structure
The `ranacapa` Shiny app requires two input files. The first requirement is a site-by-species matrix, uploaded either as a rich, dense `.biom` table, or as a tab-separated `.txt` file. Qiime2-generated`.qza` files generated by QIIME2 are not immediately suitable for ranacapa, as they do not contain full taxonomy information. If the site-by-species matrix is uploaded as a `.txt` file, the file should match the specifications of the output files from the Anacapa eDNA sequence analysis pipeline (). Specifically, each row in the `.txt` file must represent a taxonomic identification , and each column (save one) must represent the number of times that ASV appears in each sequenced sample. One column, named `sum.taxonomy` must contain the taxonomic identification, with taxonomic rank separated by a semicolon, as follows:

```
                                                                        sum.taxonomy Arch_point_1 Arch_point_2 Black_seabass_reef_1 Black_seabass_reef_2
  Annelida;Clitellata;Haplotaxida;Hormogastridae;Hormogaster;Hormogaster castillana            0            0                    0                    0
 Annelida;Clitellata;Haplotaxida;Lumbricidae;Allolobophora;Allolobophora chlorotica            0            0                    0                    0
   Annelida;Clitellata;Haplotaxida;Megascolecidae;Amynthas;Amynthas sp. YN201102-05            0            0                    0                    0
        Annelida;Clitellata;Haplotaxida;Megascolecidae;Amynthas;Amynthas thakhantho            0            0                    0                    0
       Annelida;Polychaeta;Capitellida;Maldanidae;Axiothella;Axiothella rubrocincta            0            0                    0                    0
```

The second requirement is a tab-separated `.txt` file that contains sample metadata. The first column should match the sample names in the taxonomy table; the remaining columns of the metadata file contain sample information for each of the samples in the site-by-species matrix. The metadata should contain categorical variables with two or more categories per variable. A valid metadata file for the taxonomy table above would be as follows:

```
                Sample Sample_or_Control         Island  Protection  SST
         Arch_point_1            Sample Santa Barabara unprotected   17
         Arch_point_2            Sample Santa Barabara unprotected   17
 Black_seabass_reef_1            Sample        Anacapa         MPA   16
 Black_seabass_reef_2            Sample        Anacapa         MPA   16
```

@@

## Use Cases
We expect that researchers with bioinformatic expertise will use the pipeline of their choice to assign taxonomy to eDNA datasets and generate clean taxonomy and metadata files that can be visualized in `ranacapa`. Researchers can share these files with their partners, and emphasize the analyses or visualizations most appropriate to their use case. We now show how `ranacapa` can facilitate authentic communication between researchers and community partners in two settings.

### Use Case 1: How `ranacapa` facilitated a collaboration between eDNA researchers and managers at the National Park Service

We expect the interactive taxonomy heatmap to be especially useful for researchers collaborating with natural resource managers whose efforts are focused a targetted list of rare or invasive taxa. For example, UCLA researchers who partner with resource managers at the Channel Islands National Park Service to assess the potential for eDNA as a biodiversity monitoring tool to supplement expensive and time-intensive visual biodiversity surveys in the Southern California Channel Islands [@Lessios_1996, @Usseglio_2015; @Deiner_2017] use ranacapa to share eDNA sequencing results. For this partnership, resource managers collected and filtered thirty-1L water samples for eDNA analysis at permanent monitoring sites inside and adjacent to protected areas, and research scientists at UCLA performed eDNA sequencing of the mitochondrial 12S and CO1 gene, targeting bony fishes, elasmobranches, and invertebrate taxa. The researchers processed sequences and assigned taxonomy using the Anacapa toolkit, and shared results with the resource managers using the `ranacapa` Shiny app.

@@

### Use Case 2: How `ranacapa` helps undergraduate environmental microbiology students pursue sophisticated microbiome analyses

We expect `ranacapa`'s exploratory visualizations and the accompanying explanations to help researchers introducing ecology concepts and analyses in classrooms or other informal education settings. For example, `ranacapa` was used by a research-based environmental microbiology course at UCLA [@Shapiro_2015], in which students used eDNA metabarcoding to study the impact of a local wildfire on the plant and soil microbial community. The goal of this twenty-week course was to provide undergraduate students an authentic experience in basic microbiology and microbial community ecology research. Over the first ten weeks, the instructional team, which included eDNA research scientists, extracted total DNA from student-collected soil samples and sequenced the ITS2 [@Gu_2013] and 16S SSU RNA [@Caporaso_2012] metabarcoding region to characterize the plant, bacterial, and archaeal communities. The instructional team processed sequences and assigned taxonomy using the Anacapa toolkit.

Students were introduced to the structure of eDNA sequencing results and were taught to explore data and perform simple statistical analyses using `ranacapa`. A key benefit of using `ranacapa` was that despite having no prior bioinformatics experience, students could begin exploring the biodiversity in their samples in a matter of minutes by using the online instance of Shiny app. This allowed the instructors to focus classroom time on biological questions rather than on troubleshooting bioinformatics problems, as had been the case in previous sessions of the course. The course instructors noted that visualizing eDNA data in `ranacapa` helped students understand the relationships between community profile and the various metadata they had collected in the field. By significantly reducing the time and difficulty in visualizing basic biodiversity patterns, `ranacapa` helped students develop and pursue more sophisticated analyses during the remainder of the course, using most sophisticated tools such as STAMP [@Parks_2014] and PICRUSt [@Langille_2013]. The taxonomy tables and metadata files used in this course are available online at XXXX.

@@

Metabarcode sequencing of environmental DNA is becoming a key tool in a wide variety of ecological studies, and results from these studies are of interest to a broad audience.  Our `R` package and Shiny web-app `ranacapa` helps users conduct exploratory analyses and visualizations on eDNA datasets, and is a step toward making data and analyses from eDNA sequencing-based studies more accessible and understandable for a wide range of community research partners.

We propose three avenues for future work in ranacapa. First, we plan to use `ranacapa` as the primary tool to present eDNA results from hundreds of samples sequenced by the CALeDNA community science program.  Second, `ranacapa` will be a key tool in the upcoming undergraduate curriculum module "Pipeline for Undergraduate Microbiome Analysis", which is being built as an open-source, comprehensive suite of analysis and data visualization tools for undergraduate researchers. Finally, in the long-term, we believe there is great promise in connecting `ranacapa` to packages that connect with APIs of online biodiversity databases (e.g. Taxize, rinat). This will help users explore a much wider range of biodiversity questions, for example, by programmatically asking whether their samples include invasive species that are absent from other nearby sites. In sum, tools like `ranacapa` that allow non-technical audiences to easily interact with results from eDNA sequencing studies have great potential to engage community partners with a wide range of backgrounds and interests in primary research.

@@

#### Figure 2
```{r, fig.cap="Figure 2: Taxonomy heatmap as shown in the `ranacapa` Shiny app. Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the heatmap."}
fig2
```
#### Figure 3
```{r, fig.cap="Figure 3: Taxonomy barplot as shown in the `ranacapa` Shiny app. Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the barplot"}
fig3
```

#### Figure 4
```{r, fig.cap="Figure 4: Alpha diveristy boxplots as shown in the `ranacapa` Shiny app. Users can select the X-axis variable using a dropdown menu in the app."}
fig4
```
#### Figure 5
```{r, fig.cap="Figure 5: PCoA ordination of the samples as shown in the `ranacapa` Shiny app. Users can select the grouping variable with a dropdown menu in the app."}
fig5
```

>>>>

---
title: 'ranacapa: An R package to explore environmental DNA data with exploratory statistics and interactive visualizations'
author: Gaurav S. Kandlikar (1,+), Zachary J. Gold (1), Madeline C. Cowen (1), Rachel  S. Meyer (1), Amanda C. Freise (2), Nathan J.B. Kraft (1), Jordan Moberg-Parker (2), Joshua Sprague (3), David Kushner (3), and Emily E. Curd (1)
fontsize: 11pt
output: 
  word_document:
    reference_docx: rendering-templates/word_template.docx
urlcolor: blue
always_allow_html: yes
bibliography: bibliography.bib
csl: rendering-templates/ecology-letters.csl
editor_options: 
  chunk_output_type: console
---

@@

The targeted amplification and sequencing of DNA that living organisms shed into their physical environment, termed "environmental DNA (eDNA) metabarcoding",  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].  Sequencing of environmental DNA extracted from field-collected soil, water, or sediment samples can yield insight into a range of questions, ranging from profiling the composition of ancient plant and animal communities [@Pedersen_2014], to motoring populations of rare or endangered species [@Balasingham_2017]. As the cost of eDNA metabarcoding declines and sample collection techniques become more streamlined (e.g. @Thomas_2018), professional research scientists are increasingly using eDNA metabarcoding as a platform to engage a diversity of community partners, including natural resource managers, undergraduate students, and citizen scientists in primary research. However, developing robust and impactful community science programs that engage community partners in all steps of the research process remains a challenge.

eDNA metabarcoding-based projects work well for programs that partner researchers with community scientists because non-experts can be quickly trained to collect samples in the field, and because eDNA metabarcoding is an exciting framework for research pertinent to disciplines such as medicine, agriculture, ecology, and geography [@Deiner_2017]. Community partners in such programs can have heterogeneous backgrounds, ranging from curious members of the public for whom collecting samples in the field is their first scientific research experience (e.g. University of California’s CALeDNA program, http://www.ucedna.com/), to professional natural resource managers who regularly collaborate with research scientists (e.g. Center for Ocean Solutions’ eDNA project, https://oceansolutions.stanford.edu/project-environmental-dna). A key ingredient to promote sustained engagement of community partners in these programs is that they should be able to participate across multiple stages of the research project, not only in sample collection [@Pandya_2012; @ECSA_2015]. This can be a challenge for eDNA metabacoding-based community science programs because although it is relatively easy to train community partners to collect eDNA samples, it is far more challenging to train them to independently visualize and analyze results from these studies. Indeed, learning the bioinformatic tools necessary for managing the large, multidimensional datasets generated in these studies can be difficult for professional researchers (Carey & Papin 2018), let alone for the non-technical audience of community science programs.

To address this challenge, we created an R package “ranacapa”, at the core of which is a Shiny webapp that can be used to visualize results from eDNA sequencing studies and perform simple community ecology analyses. `ranacapa` complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, `ranacapa` includes brief explanations of several core analyses used in eDNA studies core and includes links to additional educational resources. `ranacapa` works with community matrices generated via QIIME [Caporaso_2010] or the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), which is used extensively by the CALeDNA program.

In the remainder of this manuscript, we describe the package and how it is used by two community science partnerships based at the University of California, Los Angeles (UCLA): first, a collaboration between eDNA researchers and resource managers at the National Park Service, and second, a partnership between community ecology researchers and an undergraduate microbiology course at UCLA. As we show in the Use Cases, empowering community partners to interact with the data and perform simple but insightful community ecology analyses can help make these collaborations more enriching and valuable to both parties.

## Implementation
At the core of `ranacapa` is a Shiny webapp (@Chang_2018), which is avialable at http://gauravsk.shinyapps.io/ranacapa or with `ranacapa::runRanacapa()`. The package also includes two categories of helper functions (Table 1) that transform user-uploaded taxonomy and metadata tables into `R` objects that can be visualized and analyzed using the Phyloseq [@McMurdie_2013] and Vegan [@Oksanen_2018] packages. `ranacapa` is available for installation from Github or CRAN:

@@

- **Sequencing depth**:  Introduces the potential for variation in sequencing depth among samples and explains the basic logic behind rarefying samples in metagenomics studies (Figure 1). Users can rarefy the dataset to a sampling depth, or can proceed through the rest of the app without rarefying samples. The documentation acknowledges recent disagreement regarding the value of rarefying in metabarcoding and eDNA sequencing studies [@McMurdie_2014].

- **Taxonomy heatmap**: Shows the taxon-by-sample matrix as an interactive heatmap made using `heatmaply::heatmaply()`, where the color of each cell represents the number of times a given taxon was sequenced in a sample (Figure 2). Users can filter the taxon list by selecting or deselecting specific taxa.

@@

- **Alpha diversity plots**: Introduces the concept of alpha diversity as the local diversity measured in a single habitat or sample. Users can plot alpha diversity as observed taxon richness or as Shannon diversity per sample, or can group samples according to a variable in the metadata file (Figure 4).

- **Alpha diversity statistics**: Allows users to choose a variable from the metadata, and generates an alpha diversity ANOVA table according to the user-selected variable. The tab also shows the output from  a post-hoc Tukey test.

- **Beta diversity plots**: Introduces the concept of Beta diversity as the turnover in species composition across habitats (or samples). The tab includes an ordination plot generated by `phyloseq::plot_ordination()`, which in turn uses as ordination made with `phyloseq::ordinate(., method = "PCoA")`. Points on the PCoA plot are colored according to a user-selected metadata variable (Figure 5).

This tab also includes a dendrogram that groups sites based on Ward’s cluster analysis (`stats::hclust(distance_object, method = "ward.d2")`), where `distance_object` is made using `phyloseq::distance()`. For both figures, users can toggle between using Jaccard and Bray-Curtis dissimilarity.

- **Beta diversity statistics**: Shows results from two statistical tests of species turnover across site. The first statistical test is a multivariate ANOVA implemented with `vegan::adonis())`, which shows results from a statistical test of taxon turnover across sites. The second statistical test is of heterogeneity of variances among samples implemented with `vegan::betadisper()`, dthat compares the degree of sample-to-sample variation within habitats (or within other user-selected groups).

@@

### Input file structure
The `ranacapa` Shiny app requires two input files. The first requirement is a taxon-by-sample matrix, uploaded either as a rich, dense `.biom` table, or as a tab-separated `.txt` file. Qiime2-generated`.qza` files generated by QIIME2 are not immediately suitable for ranacapa, as they do not contain full taxonomy information. If the site-by-species matrix is uploaded as a `.txt` file, the file should match the specifications of the output files from the Anacapa eDNA sequence analysis pipeline (https://github.com/limey-bean/Anacapa). In Anacapa output, each row represents a taxonomic identification, and each column (save one) represents the number of times that taxon appears in each sequenced sample. One column, named `sum.taxonomy` must contain the taxonomic identification, with taxonomic rank separated by a semicolon. A valid input file is structured as follows:

```{r print-taxon-table-structure}
set.seed(123)
# randomly select 10 samples to show
samples <- runif(10, min = 1, max = nrow(taxDat))
taxDat[samples,1:6]
```

The second requirement is a tab-separated `.txt` file that contains sample metadata. The first column in the metadata file should match the sample names in the taxonomy table; the remaining columns contain sample information for each of the samples in the taxon-by-site matrix. The metadata should contain categorical variables with two or more categories per variable. A valid metadata file for the taxonomy table above is structured as follows:

```{r print-medata-table-structure}
metDat[which(metDat[,1] %in% colnames(taxDat)[2:6]),1:5]
```

@@

## Use Cases
We expect that researchers with bioinformatic expertise will use the sequence analysis pipeline of their choice to assign taxonomy to eDNA datasets, and generate clean taxonomy and metadata files that can be visualized in `ranacapa`. Researchers can share these files with their partners, and emphasize the analyses or visualizations most appropriate to their use case. We now show how `ranacapa` can facilitate authentic communication between researchers and community partners in two settings.

### Use Case 1: Partnership between eDNA researchers and natural resource managers

A team of UCLA researchers partnered with resource managers at the Channel Islands National Park Service to assess the potential for eDNA as a biodiversity monitoring tool to supplement expensive and time-intensive visual biodiversity surveys in the Southern California Channel Islands [@Lessios_1996; @Usseglio_2015; @Deiner_2017]. For this partnership, resource managers collected and filtered thirty-1L water samples for eDNA analysis at permanent monitoring sites inside and adjacent to protected areas, and research scientists at UCLA performed eDNA sequencing of the mitochondrial 12S and CO1 gene, targeting bony fishes, elasmobranches, and invertebrate taxa. The researchers processed sequences and assigned taxonomy using the Anacapa pipeline, and shared results with the resource managers using the `ranacapa` Shiny app.

@@

### Use Case 2: Partnership between eDNA researchers and an undergraduate microbiology course

A team of community ecology and environmental DNA researchers collaborate with instructors of a research-based environmental microbiology course at UCLA [@Shapiro_2015], in which students used eDNA metabarcoding to study the impact of a local wildfire on the plant and soil microbial community. The goal of this twenty-week course was to provide undergraduate students an authentic experience in basic microbiology and microbial community ecology research. Over the first ten weeks, eDNA researchers on the instructional team sequenced the ITS2 [@Gu_2013] and 16S SSU RNA [@Caporaso_2012] metabarcoding regions from student=collected soil samples and used the Anacapa pipeline to generate taxon-by-sample tables.

The course instructors used the `ranacapa` Shiny app to introduce students to the structure of eDNA sequencing results. The students were encouraged to explore data the statistical analyses most pertinent to the hypotheses they had formed at the beginning of the course. A key benefit of using `ranacapa` was that despite having no prior bioinformatics experience, students could begin exploring the biodiversity in their samples in a matter of minutes by using the online instance of Shiny app. This allowed the instructors to focus classroom time on biological questions rather than on troubleshooting bioinformatics problems, as had been the case in previous sessions of the course. The course instructors noted that visualizing eDNA data in `ranacapa` helped students understand the relationships between taxon-by-site matrices and the various metadata they had collected in the field. By significantly reducing the time and difficulty in visualizing basic biodiversity patterns, `ranacapa` helped students develop and pursue more sophisticated analyses during the remainder of the course, using most sophisticated tools such as STAMP [@Parks_2014] and PICRUSt [@Langille_2013]. The taxonomy tables and metadata files used in this course are available online at XXXX.

@@

Metabarcode sequencing of environmental DNA is becoming a key tool in a wide variety of ecological studies, and results from these studies are of interest to a broad audience.  Our `R` package and Shiny web-app `ranacapa` helps users conduct exploratory analyses and visualizations on eDNA datasets, and is a step toward more fully engaging participants in all phases of eDNA sequencing-based community science projects.

We propose three avenues for future work with ranacapa. First, we plan to use `ranacapa` as the primary tool to present eDNA results from hundreds of samples sequenced by the CALeDNA community science program.  Second, `ranacapa` is being integrated into the upcoming undergraduate curriculum module "Pipeline for Undergraduate Microbiome Analysis", which will be an open-source, comprehensive suite of analysis and data visualization tools for undergraduate researchers. Finally, in the long-term, we believe there is great promise in linking `ranacapa` with packages that connect with APIs of online biodiversity databases (e.g. Taxize [Chamberlain_2013], rinat [@Barve_2017]). This will help users explore a much wider range of biodiversity questions, for example, by programmatically asking whether their samples include invasive species that are absent from other nearby sites. In sum, tools like `ranacapa` that allow non-technical audiences to easily interact with results from eDNA sequencing studies have great potential to engage community partners with a wide range of backgrounds and interests in primary research.

@@

#### Figure 2
```{r, fig.cap="Figure 2: Taxonomy heatmap as shown in the `ranacapa` Shiny app. Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the heatmap. Users can also filter  select or deselect individual taxa to be shown in the heatmap."}
fig2
```
#### Figure 3
```{r, fig.cap="Figure 3: Taxonomy barplot as shown in the `ranacapa` Shiny app. Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the barplot"}
fig3
```

#### Figure 4
```{r, fig.cap="Figure 4: Alpha diversity boxplots as shown in the `ranacapa` Shiny app. Users can select the X-axis variable using a dropdown menu in the app."}
fig4
```
#### Figure 5
```{r, fig.cap="Figure 5: PCoA ordination of the samples as shown in the `ranacapa` Shiny app. Users can select the grouping variable with a dropdown menu in the app."}
fig5
```
